---
title: "02 Statistics"
subtitle: "Biotech 7005: Bioinformatics and Systems Modelling"
author: "Steve Pederson"
date: "30 July 2018"
output: 
  ioslides_presentation: 
    widescreen: yes
    css: custom.css
    fig_caption: yes
    logo: ../images/UoA_logo_col_vert.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE, warning = FALSE)
```

```{r}
library(dplyr)
library(ggplot2)
library(pander)
library(magrittr)
library(reshape2)
```


## Lecture Outline

1. Hypothesis Testing
    <!-- - The Null Hypothesis -->
    <!-- - $t$-tests -->
    <!-- - Wilcoxon Rank-Sum Tests -->
    <!-- - Enrichment Tests -->
2. Multiple Testing Considerations
    <!-- - Controlling the FWER -->
    <!-- - Controlling the FDR -->
    
# Hypothesis Testing

## Hypothesis Testing

In biological research we often ask: 

<center>**“Is something happening?”** or **“Is nothing happening?”**</center><br>

We might be comparing:

- Cell proliferation in response to antibiotics in media
- mRNA abundance in two related cell types
- Allele frequencies in two populations
- Methylation levels across genomic regions

## Hypothesis Testing {.build}

How do we decide if our experimental results are "significant"?

- Is it normal variability?
- What would the data look like if our experiment had no effect?
- What would our data look like if there was an effect?

Every experiment is considered as a **random sample** from all possible repeated experiments.
    
## Sampling | Examples 
    
Most experiments involve measuring something:

- Discrete values e.g. read counts, number of colonies
- Continuous values e.g. $C_t$ values, fluorescence intensity

Every experiment is considered as a **random sample** from all possible repeated experiments.
    
## Sampling | Examples 

Many data collections can also be considered as experimental datasets

In the 1000 Genomes Project a risk allele for T1D has a frequency of $\pi = 0.07$ in European Populations.  

- **Does this mean, the allele occurs in exactly 7% of Europeans?**

## Sampling | Examples 

In our *in vitro* experiment, we found that 90% of HeLa cells were lysed by exposure to our drug.

- **Does this mean that exactly 90% of HeLa cells will always be destroyed?**
- **What does this say about *in vivo* responses to the drug?**

## Population Parameters {.build}

- Experimentally-obtained values represent an **estimate** of the **true effect**
- More formally referred to as **population-level** parameters
- Every experiment is considered a random sample of the complete population
- Repeated experiments would give a different (but similar) estimate

All population parameters are considered to be fixed values, e.g. 

- Allele frequency ($\pi$) in a population
- The **average change** in mRNA levels 

## Hypothesis Testing | The Null Hypothesis

All classical statistical testing involves:

1) a **Null Hypothesis** ($H_0$) and 
2) an **Alternative Hypothesis** ($H_A$)

**Why do we do this?**

## Hypothesis Testing | The Null Hypothesis

- We define $H_0$ so that *we know what the data will look like if there is no effect* 
- The alternate ($H_A$) includes every other possibility besides $H_0$

An experimental hypothesis may be

$$
H_0: \mu = 0 \quad Vs \quad H_A: \mu \neq 0
$$

Where $\mu$ represents the average change in a value (e.g. mRNA expression levels)

## The Sample Mean | Normally Distributed Data

For every experiment we conduct we can get two key values:

1: The sample mean ($\bar{x}$) estimates the population-level mean (e.g. $\mu$)

$$
\text{For} \quad \mathbf{x} = (x_1, x_2, ..., x_n) \\
\bar{x} = \frac{1}{n}\sum_{i = i}^n x_i
$$

*This will be a different value every time we repeat the experiment*

## The Sample Variance | Normally Distributed Data

For every experiment we conduct we can get two key values:

2: The sample variance ($s^2$) estimates the population-level variance ($\sigma^2$)

$$
s^2 = \frac{1}{n-1} \sum_{i = 1}^n (x_i - \bar{x})^2
$$

*This will also be a different value every time we repeat the experiment*

## The Sample Mean | A qPCR Experiment

```{r, echo=FALSE}
x <- c(2.1, 2.8, 2.5, 2.6)
```

- Comparing expression levels of *FOXP3* in T~reg~ and T~h~ cells ($n = `r length(x)`$ donors)
- The change within each donor is obtained as $\Delta \Delta C_t$ values 
- $\mathbf{x} =$ (`r pander(x)`)

## The Sample Mean | A qPCR Experiment

- Comparing expression levels of *FOXP3* in T~reg~ and T~h~ cells ($n = `r length(x)`$ donors)
- The change within each donor is obtained as $\Delta \Delta C_t$ values 
- $\mathbf{x} =$ (`r pander(x)`)

<center>$H_0: \mu = 0$ Vs $H_A: \mu \neq 0$ </center><br>

where $\mu$ is the average change in *FOXP3* expression in the entire population

## The Sample Mean | A qPCR Experiment

Now we can get the sample mean:

$$
\begin{aligned}
\bar{x} &= \frac{1}{n}\sum_{i = i}^n x_i \\
        &= \frac{1}{4}(2.1 + 2.8 + 2.5 + 2.6) \\
        &= `r mean(x)`
\end{aligned}
$$

This is our estimate of the true change in expression ($\mu$)

## The Sample Mean | A qPCR Experiment
And the sample variance:

$$
\begin{aligned}
s^2 &= \frac{1}{n - 1} \sum_{i = 1}^n (x_i - \bar{x})^2\\
    & = \frac{1}{3}\sum_{i = 1}^4 (x_i - 2.5)^2\\
    &= `r sprintf("%0.4f", var(x))`
\end{aligned}
$$


## The Sample Mean

- Every time we repeat an experiment, we get a different value for $\bar{x}$
- Would these follow any specific pattern?

## The Sample Mean

- Every time we repeat an experiment, we get a different value for $\bar{x}$
- Would these follow any specific pattern?
- **They would be normally distributed around the true value!**

$$
\bar{x} \sim \mathcal{N}(\mu, \frac{\sigma}{\sqrt{n}})
$$

where:

- $\mu$ represents the true population mean
- $\sigma$ represents the variance in the population (probably unknown)

## The Null Hypothesis

If $H_0$ is true (and we know $\sigma$), **we know what the data will look like**.

- A simple transformation gives us a $Z$ statistic where $Z \sim \mathcal{N}(0, 1)$

$$
Z = \frac{\bar{x} - \mu}{\sigma / \sqrt{n}}
$$

- If  $H_0$ is true $Z$ will come from $\mathcal{N}(0, 1)$
- If $H_0$ is NOT true $Z$ will come from some other distribution
- We compare our results (i.e. $Z$) to $\mathcal{N}(0, 1)$ and see if our results are likely or unlikely


## The Null Hypothesis

```{r}
df <- data_frame(x = seq(-5, 5, length.out = 1000),
           y = dnorm(x)) 
```


**If $H_0$ is true, where would we expect $Z$ to be?**  
**If $H_0$ is NOT true, where would we expect $Z$ to be?**

```{r, fig.show='asis', fig.align='center', fig.height=4}
ggplot(df, aes(x = x, y = y)) +
  geom_path() +
  theme_bw() +
  labs(x = "Z", y = "") +
  scale_y_continuous(expand = c(0, 0)) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        axis.line.x = element_line(colour = "black"),
        panel.border = element_blank(),
        panel.grid = element_blank())
```

## The Null Hypothesis

**Would a value $Z > 1$ be likely if $H_0$ is TRUE?**

```{r, echo=FALSE, fig.show='asis', fig.align='center', fig.height=4}
ggplot(df, aes(x = x, y = y)) +
  geom_path() +
  geom_ribbon(data = df[df$x >1, ],
              aes(x= x, ymax =y, ymin = 0),
              fill = "grey") +
  geom_vline(xintercept = 1) +
  theme_bw() +
  labs(x = "Z", y = "") +
  scale_y_continuous(expand = c(0, 0)) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        axis.line.x = element_line(colour = "black"),
        panel.border = element_blank(),
        panel.grid = element_blank())
```

## The Null Hypothesis

**Would a value $>2$ be likely if $H_0$ is `TRUE`?**

```{r, echo=FALSE, fig.show='asis', fig.align='center', fig.height=4}
ggplot(df, aes(x = x, y = y)) +
  geom_path() +
  geom_ribbon(data = df[df$x >2, ],
              aes(x= x, ymax =y, ymin = 0),
              fill = "grey") +
  geom_vline(xintercept = 2) +
  theme_bw() +
  labs(x = "Z", y = "") +
  scale_y_continuous(expand = c(0, 0)) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        axis.line.x = element_line(colour = "black"),
        panel.border = element_blank(),
        panel.grid = element_blank())
```



## $p$ Values

- The area under all probability distributions adds up 1
- The area to the right of 2, is the probability of obtaining $Z >2$
- This is `r sprintf("%.3f", 1- pnorm(2))`
- Thus **if $H_0$ is true, we know the probability of obtaining a $Z$-statistic $>2$**

## $p$ Values

**In our qPCR experiment, could the $\Delta \Delta C_t$ values be either side of zero?**

- This means we also need to check the values for the other extreme
- This distribution is symmetric around zero:
    + $p(|Z| > 2)$ = $p(Z > 2) + P(Z < -2)$ 
- Known as a *two-sided* test    

**This is the most common way of deciding if $H_0$ is true!**

## $p$ Values

```{r, echo=FALSE, fig.show='asis', fig.align='center', fig.height=4}
ggplot(df, aes(x = x, y = y)) +
  geom_path() +
  geom_ribbon(data = df[df$x >2, ],
              aes(x= x, ymax =y, ymin = 0),
              fill = "grey") +
  geom_ribbon(data = df[df$x < -2, ],
              aes(x= x, ymax =y, ymin = 0),
              fill = "grey") +
  geom_vline(xintercept = c(-2, 2)) +
  theme_bw() +
  labs(x = "Z", y = "") +
  scale_y_continuous(expand = c(0, 0)) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        axis.line.x = element_line(colour = "black"),
        panel.border = element_blank(),
        panel.grid = element_blank())
```


## $p$ Values

$$
Z = \frac{\bar{x} - \mu}{\sigma / \sqrt{n}}
$$

1. We first calculate $Z$, and compare to $\mathcal{N}(0, 1)$ 
2. We obtain the probability of obtaining a $Z$-statistic *at least as extreme* as our value
    + If $H_0$ is true, $Z$ will come from $\mathcal{N}(0, 1)$
    + If $H_0$ is NOT true, we have no idea where $Z$ will be. 
    + It can be anywhere $-\infty < Z < \infty$
    
## $p$ Values

### Definition

> A $p$ value is the probability of observing data as extreme, or more extreme than we have observed, if $H_0$ is true.

To summarise:

1. We calculate a test statistic $Z$ using $\mu = 0$ and $\sigma$
2. Compare this to $\mathcal{N}(0, 1)$ to find the probability ($p$) of observing data as extreme, or more extreme, than we observed if $H_0$ is true
3. If $p$ is low (e.g. $p<0.05$), we reject $H_0$ as unlikely and accept $H_A$

## $t$-tests

In reality, we will never know the population variance ($\sigma^2$), just like we will never know $\mu$

- If we knew these values we wouldn't need to do any experiments
- We can use our *sample variance* ($s^2$) to estimate $\sigma^2$

**Due to the uncertainty introduced by using $s^2$ instead of $\sigma^2$ we can no longer compare to the $Z \sim \mathcal{N}(0, 1)$ distribution.**

## $t$-tests

Instead we use a $T$-statistic

$$
T = \frac{\bar{x} - \mu}{s / \sqrt{n}}
$$

Then we compare to a $t$-distribution

## $t$-tests | The t distribution

The $t$-distribution is very similar to $\mathcal{N}(0, 1)$

- Bell-shaped & symmetrical around zero
- Has fatter tails $\implies$ extreme values are more likely
- The parameter *degrees of freedom* (df) specifies how "fat" the tails are
- As df $\rightarrow 0$ the tails get fatter


## $t$-tests | The t distribution

```{r, fig.show='asis', fig.align='center', fig.height=4.5}
df %>% 
  select(x) %>%
  mutate(Z = dnorm(x),
         t_3 = dt(x, df = 3),
         t_10 = dt(x, df = 10)) %>%
  melt(id.vars = "x") %>%
  mutate(variable = factor(variable, levels = c("t_3", "t_10", "Z"))) %>%
  ggplot(aes(x = x, y= value, colour = variable)) +
  geom_line(size = 1)+
  theme_bw() +
  labs(x = "", y = "",colour = "Distribution") +
  scale_y_continuous(expand = c(0, 0)) +
  scale_colour_discrete(breaks =  c("t_3", "t_10", "Z"),
                        labels = c(expression(t[3]), expression(t[10]), expression(Z))) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        axis.line.x = element_line(colour = "black"),
        panel.border = element_blank(),
        panel.grid = element_blank(),
        legend.position = c(0.85, 0.7),
        legend.text = element_text(size = 11))
```

## $t$-tests | Degrees Of Freedom

At their simplest:

$$
df = n - 1
$$

As $n$ increases $s^2 \rightarrow \sigma^2$ and $\implies t_{df} \rightarrow Z$

<!-- More samples:   -->
<!-- $\implies$ less chance of an extreme value under $H_0$   -->
<!-- $\implies$ more statistical power -->

##  $t$-tests | qPCR Data

For our qPCR data, testing $\mu = 0$:

$$
\begin{aligned}
T &= \frac{\bar{x} - \mu}{s / \sqrt{n}} \\
  &= \frac{`r mean(x)` - 0}{`r sd(x)` / \sqrt{`r length(x)`}} \\
  &= `r sprintf("%.3f", mean(x) / (sd(x) / sqrt(length(x))))`
\end{aligned}
$$

##  $t$-tests | qPCR Data

```{r, fig.show='asis', fig.align='center', fig.height=4.5}
t <- t.test(x)$statistic
df %>% 
  select(x) %>%
  mutate(Z = dnorm(x),
         t_3 = dt(x, df = 3),
         t_10 = dt(x, df = 10)) %>%
  melt(id.vars = "x") %>%
  mutate(variable = factor(variable, levels = c("t_3", "t_10", "Z"))) %>%
  ggplot(aes(x = x, y= value, colour = variable)) +
  geom_line(size = 1)+
  geom_vline(xintercept = t) +
  theme_bw() +
  labs(x = "", y = "",colour = "Distribution") +
  scale_y_continuous(expand = c(0, 0)) +
  scale_colour_discrete(breaks =  c("t_3", "t_10", "Z"),
                        labels = c(expression(t[3]), expression(t[10]), expression(Z))) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        axis.line.x = element_line(colour = "black"),
        panel.border = element_blank(),
        panel.grid = element_blank(),
        legend.position = c(0.85, 0.7),
        legend.text = element_text(size = 11))
```

##  $t$-tests | qPCR Data

- Our degrees of freedom here are $\nu = 3$
- By comparing to $t_3$ we get

$$p = `r sprintf("%0.5f", t.test(x)[["p.value"]])`$$

- Thus, the probability of observing data as (or more) extreme as this is very low if $H_0$ was true (1 in ~2300 experiments).
- We reject $H_0$ and accept $H_A$

## Hypothesis Testing | Summary

1. Define $H_0$ and $H_A$
2. Calculate sample mean ($\bar{x}$) and variance ($s^2$)
3. Calculate $T$-statistic and degrees of freedom ($\nu$)
4. Compare to $t_{\nu}$ and obtain probability of observing $\bar{x}$ if $H_0$ is true
5. Accept or reject $H_0$ if $p < 0.05$ (or some other value)

## Hypothesis Testing | Summary

1. Define $H_0$ and $H_A$
2. Calculate sample mean ($\bar{x}$) and variance ($s^2$)
3. Calculate $T$-statistic and degrees of freedom ($\nu$)
4. Compare to $t_{\nu}$ and obtain probability of observing $\bar{x}$ if $H_0$ is true
5. Accept or reject $H_0$ if $p < 0.05$ (or some other value)

- This applies to most situations
- Assumes data is normally distributed

## Two Sample $t$-tests {.build}

In the above we had the $\Delta \Delta C_t$ values within each donor.

**What if we just had 4 values from each cell-type from different donors?**

- We would be interested in the difference between the two means ($\mu_A$ and $\mu_B$)
- We could use a two sample $t$-test to compare $\bar{x}_A$ and $\bar{x}_B$
- The principle is the same, calculations are different

## Two Sample $t$-tests 

For $H_0: \mu_A = \mu_B$ Vs $H_A: \mu_A \neq \mu_B$

1. Calculate the two sample means: $\bar{x}_A$ and $\bar{x}_B$
2. Calculate the two sample variances $s_A^2$ and $s_B^2$
3. Calculate the pooled denominator: $\text{SE}_{\bar{x}_A - \bar{x}_B}$
    - Formula varies for equal/unequal variances
4. Calculate the degrees of freedom
    - Formula varies for equal/unequal variances
    
## Two Sample $t$-tests 

$$
T = \frac{\bar{x}_A - \bar{x}_B}{\text{SE}_{\bar{x}_A - \bar{x}_B}}
$$

If $H_0$ is true then

$$
T \sim t_{\nu}
$$

1. We  compare our test-statistic to this distribution
2. Are we likely to see this value (or more extreme) under $H_0$?
3. Accept or Reject $H_0$

# Hypothesis Testing For Non-Normal Data

## What if our data is not Normally Distributed {.build}

**When would data not be Normally Distributed?**

- Counts: These are discrete whilst normal data is continuous
- Proportions: These are bound at 0 & 1, i.e. $0 < \pi < 1$
- Data generated by Exponential, Uniform, Binomial etc. processes

Two useful tests:

- Wilcoxon Rank-Sum Test
- Fisher's Exact Test

## Wilcoxon Rank-Sum Test

$H_0: \mu_A = \mu_B$ Vs $H_A: \mu_A \neq \mu_B$

- Used for any two measurement groups which are not Normally Distributed.
- Assigns each measurement a rank
- Compares ranks between groups
- Determines probability of observing differences in ranks
- Also known as the Mann-Whitney Test
- Requires higher sample sizes than $t$-tests

## Fisher's Exact Test

- Used for $2 \times 2$ tables (or $m \times n$) with counts and categories
- Analogous to a Chi-squared ($\chi^2$) test but more robust to small values
- Commonly used to test for enrichment of an event within one group above another group (GO terms; TF motifs; SNP Frequencies)

## Fisher's Exact Test

An example table

```{r}
x <-matrix(c(12, 20, 12, 4), ncol = 2) %>%
  set_rownames(c("Upper Lakes", "Lower Plains")) %>%
  set_colnames(c("A", "B"))
pander(x)
```

<br>$H_0:$ **No association** between allele frequencies and location  
$H_A:$ There **is an association** between between allele frequencies and location

## Fisher's Exact Test

- We find the probability of obtaining tables with more extreme distributions, holding row and column totals fixed
- Can also be defined using the hypergeometric distribution


# Error Types and Multiple Hypothesis Testing 

## Rejection of $H_0$

> A $p$ value is the probability of observing data as (or more) extreme if $H_0$ is true.

We commonly reject $H_0$ if $p < 0.05$

**How often would we incorrectly reject $H_0$?**

## Rejection of $H_0$

> A $p$ value is the probability of observing data as (or more) extreme if $H_0$ is true.

We commonly reject $H_0$ if $p < 0.05$

**How often would we incorrectly reject $H_0$?**

About 1 in 20 times, we will see $p < 0.05$ if $H_0$ is true

## Error Types

- Type I errors are when we reject $H_0$ but $H_0$ is true
- Type II errors are when we accept $H_0$ when $H_0$ is false

|                    | $H_0$ `TRUE` | $H_0$ `FALSE` |
| ------------------:| ------------ | ------------- |
| Reject $H_0$       | Type I Error | $\checkmark$  |
| Don't Reject $H_0$ | $\checkmark$ | Type II Error |

**What are the consequences of each type of error?**

## Error Types

**What are the consequences of each type of error?**

Type I: Waste \$\$$ chasing dead ends  
Type II: We miss a key discovery

- In research, we usually try to minimise Type I Errors
- Increasing sample-size reduces Type II Errors

## Family Wise Error Rates

1. Imagine we are examining every human gene ($m=$ 25,000) for differential expression using RNASeq
2. Imagine there are 1000 genes which are truly DE

**How many times would we incorrectly reject $H_0$ using $p < 0.05$**

## Family Wise Error Rates

1. Imagine we are examining every gene ($m=$ 25,000) for differential expression using RNASeq
2. Imagine there are 1000 genes which are truly DE

**How many times would we incorrectly reject $H_0$ using $p < 0.05$**

We effectively have 25,000 tests, with 24,000 times $H_0$ is true

$\frac{25000 - 1000}{20} = 1200$ times

**Could this lead to any research dead-ends?**

## Family Wise Error Rates

This is an example of the Family-Wise Error Rate (i.e. Experiment-Wise Error Rate)

### Definition

> The Family-Wise Error Rate (FWER) is the probability of making one or more false rejections of $H_0$

In our example, the FWER $\approx 1$

## Family Wise Error Rates

What about if we lowered the rejection value to $\alpha = 0.001$?

We would incorrectly reject $H_0$ once in every 1,000 times

$\frac{25000 - 1000}{1000} = 24$ times

The FWER is still $\approx 1$

## The Bonferroni Adjusment

- If we set the rejection value to $\alpha* = \frac{\alpha}{m}$ we control the FWER at the level $\alpha$
- To ensure that $p$(one or more Type I errors) = 0.05 in our example:<br><br>$\implies$ Reject $H_0$ if $p < \frac{0.05}{25000}$

**What are the consequences of this?**

## The Bonferroni Adjusment

- If we set the rejection value to $\alpha* = \frac{\alpha}{m}$ we control the FWER at the level $\alpha$
- To ensure that $p$(one or more Type I errors) = 0.05 in our example:<br><br>$\implies$ Reject $H_0$ if $p < \frac{0.05}{25000}$

**What are the consequences of this?**

- Large increase in Type II Errors
- BUT what we find we are very confident about $\implies$ we don't waste time and money on dead ends!

## The False Discovery Rate

- An alternative is to allow a small number of Type I Errors in our results $\implies$ we have a False Discovery Rate (FDR)
- Instead of controlling the FWER at $\alpha = 0.05$, if we control the FDR at $\alpha = 0.05$ we allow up to 5% of our list to be Type I Errors

Most common procedure is the Benjamini-Hochberg

**What advantage would this offer?**

## The False Discovery Rate

- An alternative is to allow a small number of Type I Errors in our results $\implies$ we have a False Discovery Rate (FDR)
- Instead of controlling the FWER at $\alpha = 0.05$, if we control the FDR at $\alpha = 0.05$ we allow up to 5% of our list to be Type I Errors

Most common procedure is the Benjamini-Hochberg

**What advantage would this offer?**

- Lower Type II Errors
- 5% chance we chase a dead end

## The False Discovery Rate

For those interested, the BH procedure for $m$ tests is (not-examinable)

1. Arrange $p$-values in ascending order $p_{(1)}, p_{(2)}, ..., p_{(m)}$ 
2. Find the largest number $k$ such that $p_{(k)} \leq \frac{k}{m}\alpha$
3. Reject $H_0$ for all $H_{(i)}$ where $i \leq k$
