---
title: "Topic 3: Plotting Data"
subtitle: "Biotech 7005: Bioinformatics and Systems Modelling"
author: "Steve Pederson"
date: "30 July 2018"
output: 
  html_document: 
    fig_caption: yes
    toc: yes
    toc_depth: 2
---

```{r, loadPackages, echo = FALSE, include = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, include = TRUE, 
               warning = FALSE, message = FALSE, fig.align = "center",
               results = 'hide', fig.show = 'hide')
```

# Data Visualisation Using the Package `ggplot2`

## The package `graphics`

`R` has numerous plotting functions in the base package `graphics`

```{r, eval = FALSE}
?plot
?boxplot
?hist
```

__Go to the `Examples` at the bottom of each help page and copy a few lines__

The package `ggplot2` gives much more flexibility and power.
However, it has a unique syntax and approach where we plot by adding layers of relevant information, based on the concepts outlined in [The Grammar of Graphics](https://www.springer.com/gp/book/9780387245447).

## The package `ggplot2` 

First we need to load the package, which is part of the core `tidyverse` packages.

```{r}
library(tidyverse)
```

We'll also use the transport dataset from our previous session.
This time, we'll just overwrite the `height` column with the same values in *m* instead of *cm*.
We'll also remove `X1` and add the BMI information directly as we're loading the data.
Hopefully you can see the usefulness of the `%>%` from this example.

**Remember to load all packages at the top of your `R` script for easier reference in the future!**

```{r}
csvFile <- file.path("data", "transport.csv")
file.exists(csvFile)
data <- read_csv(csvFile) %>%
  mutate(height = height/100, 
         BMI = weight / height^2) %>%
  select(-X1)
data
```

The main function is `ggplot()`.
In this first stage we first define the aesthetic mappings using `aes()`.
These simply define what is plotted on which axis, and what defines the colour/shape etc.
When calling this function alone no data will be plotted, and we get the plot area only.

```{r, fig.show='hide'}
ggplot(data, aes(x = weight, y = height))
```

### Adding geometry

After defining the aesthetic mappings, we:

1. Tell R that "*more is to come*" by adding a `+` symbol at the end of the line
2. Add the geometry using various `geom_...()` functions

Let's start by plotting points, for which we need `geom_point()`.

```{r}
ggplot(data, aes(x = weight, y = height)) +
  geom_point()
```

There are numerous aesthetics specifically available for `geom_point()`, so let's check the help page.

```{r, eval=FALSE}
?geom_point
```

We can colour points, so let's do this based on the `transport` column.
Note that if we add this during the call to `ggplot()` this will be passed down to every subsequent `geom_*` that we call.

```{r, fig.show='hide'}
ggplot(data, aes(x = weight, y = height, colour = transport)) +
  geom_point()
```

We can also specify the shape to represent another variable, such as `gender`.

```{r}
ggplot(data, aes(x = weight, y = height, colour = transport, shape = gender)) +
  geom_point()
```

Let's leave the general aesthetics in `ggplot()`, with the colour and the shape aesthetics specifically assigned to the points.

```{r}
ggplot(data, aes(x = weight, y = height)) +
  geom_point(aes(colour = transport, shape = gender))
```

Anything set outside of the `aes()` function are general across all points.
Here we can make all points larger.
Size is a relative scale and doesn't specifically represent a value in pixels, mm or anything similar.

```{r}
ggplot(data, aes(x = weight, y = height)) +
  geom_point(aes(colour = transport, shape = gender), size = 3)
```

### Layering Multiple Geometry Types

In addition to the points, we can easily add a smoothed line.
This defaults to a loess fit, so let's select a regression line straight away by setting the `method` argument to `"lm"`.

```{r}
ggplot(data, aes(x = weight, y = height)) +
  geom_point(aes(colour = transport, shape = gender)) +
  geom_smooth(method = "lm")
```

The standard errors for the line are represented by shading by default, and we can remove these by setting `se = FALSE`.

```{r}
ggplot(data, aes(x = weight, y = height)) +
  geom_point(aes(colour = transport, shape = gender)) +
  geom_smooth(method = "lm", se= FALSE)
```


### Axis and Legend Labels

Axis and legend labels can be added using `labs()`.
The default values are the column names, so we can tidy those up fairly easily.

```{r}
ggplot(data, aes(x = weight, y = height)) +
  geom_point(aes(colour = transport, shape = gender)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Weight (kg)", y = "Height (cm)", 
       shape = "Gender", colour = "Transport")
```


### Using Facets

One of the most useful features of `ggplot2` is the ability to create multiple sub-plots using the function `facet_wrap()`.
Let's create separate plots based on the `gender` column.
This will also plot separate regression lines.

```{r}
ggplot(data, aes(x = weight, y = height)) +
  geom_point(aes(colour = transport, shape = gender)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Weight (kg)", y = "Height (cm)", 
       shape = "Gender", colour = "Transport") +
  facet_wrap(~gender) 
```

### Different geoms

Enter `geom_` in the Console followed by the `tab` key.
Numerous options will appear such as `geom_density()` and `geom_boxplot()`.
You may need to think carefully about what is plotted on which axis for some of these so pay careful attention.

Here's `geom_density()`

```{r}
ggplot(data, aes(x = height, fill = gender)) +
  geom_density(alpha = 0.5)
```

We can easily make boxplots.

```{r}
ggplot(data, aes(x = gender, y =height, fill = gender)) +
  geom_boxplot()
```

We could plot separate facets based on the transport method as well.

```{r}
ggplot(data, aes(x = gender, y =height, fill = gender)) +
  geom_boxplot() +
  facet_wrap(~transport)
```

Histograms use `geom_histogram()`

```{r}
ggplot(data, aes(x = height)) +
  geom_histogram(bins = 20)
```

These tend to need a little tweaking though.

```{r}
ggplot(data, aes(x = height)) +
  geom_histogram(bins = 20, fill = "grey50", colour = "black")
```


### Combining With `dplyr`

We can use `dplyr` (or any other) function to summarise our data before plotting, then _pipe_ (`%>%`) into `ggplot()` without modifying our original data object.
As we saw last time, this is loaded by default when we called `library(tidyverse)`.

Let's start working towards a barplot with error bars, which will use `geom_bar()`.
This has a slightly unusual set of default arguments, so we need to provide the argument `stat = "identity"`, which tells it to use the values provided and not to try summing anything.

```{r}
data %>%
  group_by(transport, gender) %>%
  summarise(mn_height = mean(height), sd_height = sd(height)) %>%
  ggplot(aes(x = transport, y = mn_height, fill = transport)) +
  geom_bar(stat = "identity") +
  facet_wrap(~gender) +
  guides(fill =FALSE)
```

To add error bars we now just use `geom_errorbar()`

```{r}
data %>%
  group_by(transport, gender) %>%
  summarise(mn_height = mean(height), sd_height = sd(height)) %>%
  ggplot(aes(x = transport, y = mn_height, fill = transport)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mn_height - sd_height,
                    ymax = mn_height + sd_height),
                width = 0.6)+
  facet_wrap(~gender) +
  guides(fill =FALSE)
```

### Setting Themes

`ggplot2` uses the idea of themes to control the overall appearance.
The default has white grid-lines and a grey background, with a good alternative being `theme_bw()`.

```{r}
ggplot(data, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw()
```

Try some of the alternatives like `theme_void()` and `theme_classic()`

The `theme()` function is also where you set `axis.text`, `legend` and a whole raft of other attributes.
The syntax often uses _elements_ to set an attribute, which can take a while to come to grips with.
Let's check the help page.

```{r}
?theme
```

As an example, changing text within themes uses `element_text()`

```{r}
?element_text
```

Let's change the text on the x-axis of our last plot.
Hopefully this gives you enough ideas on how to customise your own plots.

```{r}
ggplot(data, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw() +
  theme(axis.text.x = element_text(family = "Times", size = 15, angle = 90, colour = "blue"))
```

To remove all attributes use `element_blank()` in place of `element_rect()`, `element_text()` or `element_line()`.
In the following, we'll remove the grid-lines.

```{r}
ggplot(data, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw() +
  theme(panel.grid = element_blank())
```

### Legend Positions

We can also move the legend to multiple places:

```{r}
ggplot(data, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw() +
  theme(legend.position = "bottom")
```

Or we can use co-ordinates to place it inside the plotting region.
This method treats the plot area as a grid of `width = height = 1`.

```{r}
ggplot(data, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw() +
  theme(legend.position = c(0.85, 0.15))
```

### Colours and Axes

The scale of axes can also be set manually to be reverse or on the log10 scale.
**Note that this code has switched to a plot of weight not height.**

```{r}
ggplot(data, aes(x = gender, y = weight, fill = gender)) +
  geom_boxplot() +
  theme_bw() +
  theme(legend.position = c(0.85, 0.15)) +
  scale_y_log10(breaks = c(50, 60, 70, 80, 90)) +
  scale_fill_manual(values = c("grey70", "lightblue"))
```

In the last line of the above, we've also set the colour manually.
Once you get this plotting approach, it's very easy to customise and make great looking plots.
