---
title: "R Setup"
author: "Steve Pederson"
date: "23 July 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation Instructions {.tabset}

There are two primary pieces of software we'll need to install in preparation for the course: `R` and `RStudio`.
However, there are some differences in installation across Mac (OSX) and Windows, so please check the appropriate section below.
You will need to be connected to the internet during installation.

### OSX

1. **The [R Software](https://cran.r-project.org/bin/macosx/) itself.**
    + The latest release is 3.5.1, however any version >3.5 will be suitable. 
    + If you have Mavericks or lower (OSX $\leq$ 10.9) you will be restricted to R 3.3.3. Whilst this will be mostly compatabile with today's material, you may like to consider upgrading your OS relatively soon. Bioinformatics tools are often developed quickly and may render this OS as more and more of a problem as time moves on.
2. **The IDE [R Studio](https://www.rstudio.com/products/rstudio/download/#download)** which is the software we use to interact with R
3. R is also dependent on [XQuartz](https://www.xquartz.org/) so please ensure you have the latest version

### Windows 

If you are running your own laptop where *you are the administrator*, installation is relatively easy.
Please install the following software in order:

1. **The [R Software](https://cran.r-project.org/bin/windows/) itself.** The latest release is 3.5.1, however any version >3.5 will be suitable
2. The IDE [R Studio](https://www.rstudio.com/products/rstudio/download/#download) which is the software we use to interact with R

### Ubuntu

Open a terminal (`Ctrl + Alt +T`).
For Ubuntu 16.04 simply enter the following, however if you are on another version you'll need to change "xenial" to your specific version (i.e. trusty, bionic etc)

```
sudo echo "deb http://cran.rstudio.com/bin/linux/ubuntu xenial-cran35/" | sudo tee -a /etc/apt/sources.list
```

From there just enter the following. One line at a time is best as you may be required to enter `yes` or `no` at various times.
First we'll need to setup a key server

```
gpg --keyserver keyserver.ubuntu.com --recv-key E084DAB9
gpg -a --export E084DAB9 | sudo apt-key add -
```
From here we can install all the software we need

```
sudo apt-get update
sudo apt-get install r-base r-base-dev gdebi-core
```

For Ubuntu 16.04 or later

```
wget https://download1.rstudio.org/rstudio-xenial-1.1.453-amd64.deb
sudo gdebi -n rstudio-xenial-1.1.453-amd64.deb
rm rstudio-xenial-1.1.453-amd64.deb
```

For Ubuntu 15.10 or earlier

```
wget https://download1.rstudio.org/rstudio-1.1.453-amd64.deb
sudo gdebi -n rstudio-1.1.453-amd64.deb
rm rstudio-1.1.453-amd64.deb
```

##

Once you have this all installed, open `RStudio` and look for the frame which is titled **`Console`**. 
Paste the following lines all at once, followed by **Enter**. If you are prompted to update any packages, type  `a` for all.

```{r, eval=FALSE}
pkgs <- c("tidyverse", "pander", "plotly", "ggrepel", "gridExtra", 
          "knitr", "rmarkdown", "readxl", "devtools")
install.packages(pkgs, dependencies = TRUE )
```

This will install the required packages for the course.
These may take a while to install.
If you see an error which indicates a failure at the end, please ask a demonstrator to help

[Back to main page](./)