---
title: 'Biotech 7005: Statistics Tutorial'
author: "Steve Pederson"
date: "31 July 2018"
output: 
  html_document: 
    fig_caption: yes
    fig_width: 8
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "hide",
											fig.show = "hide",
											message = FALSE, warning = FALSE)
library(pander)
```


# Single Hypothesis Testing

## Recap

In the lectures we discussed a few things:

### Normally Distributed Data

1. Every experiment is a random sample of the complete set of all possible samples
2. We define a *null hypothesis* ($H_0$) so we can describe what our data will look like if there are **no effects due to our experiment**
3. The sample mean ($\bar{x}$) is Normally Distributed around the population mean ($\mu$)
4. We can use our sample variance ($s^2$) to estimate population variance ($\sigma^2$)
5. A $T$-statistic can be calculated using the formula <br> <center>$T = \frac{\bar{x} - \mu}{s / \sqrt{n}}$</center>
6. We compare our $T$-statistic to a $t$-distribution to see how likely more extreme values are if $H_0$ is true, and reject $H_0$ if these $p$-values are suitably low.

### Non-parametric Tests

For data which is not normally distributed we learned two new tests besides a $t$-test

1. The Wilcoxon Rank Sum Test
2. Fisher's Exact Test

## Exercises

### Q1

In the following, try to describe the value that the experiment is investigating

a. The SNP at http://exac.broadinstitute.org/variant/3-8787263-G-A was found in 0.09% of 16366 South Asian genomes, and 0.05% of 8644 East Asian genomes. 
    i. *What do these numbers represent?*
    ii. *Do you think that these values represent normally distributed data?*
b. In a qPCR experiment, the gene CTLA4 was found to be expressed in Treg cells at levels 10 times higher than those found in Th cells. *What is this number an estimate of?*
c. In three transfection experiments, the average transfection efficiency was found to be 6%. *What is this number estimating?*

### Q2

The average age of all students at the University of Adelaide was calculated. *Is this an estimate? Explain your reasoning*


### Q3

a) We think that a gene will be silenced by over-expression of a targeted siRNA. *What would our null hypothesis and alternate hypothesis be?*
b) We suspect that a given SNP will provide increased zebrafish survival in a low-oxygen environment. We place 100 fish in a low-oxygen tank, and 100 fish in a normal-oxygen tank for 6 hours. At the end, we collect the survivors in each tank and genotype them at this SNP, giving the following results:

|          | Allele A | Allele B |
|:-------- | --------:| --------:|
| Hypoxia  | 10       |      32  |
| Normoxia | 94       |      86  |

*What would the null and alternate hypotheses be for this experiment?*


### Q4

`R` allows us to easily take random samples from a normal distribution.
We can generate 10 random samples from $\mathcal{N}(5, 1)$ using the following code.

```{r}
x <- rnorm(n = 10, mean = 5, sd = 1)
```

We could then find the mean of our random sample using the following

```{r}
mean(x)
```

a) *Why was this value not equal to 5?*
b) *Repeat the random sampling several times. Do you ever get 5 exactly?*

We could actually repeat this 1000 times, very quickly in `R`.
The following command will form a matrix with 1000 columns, and each column will be the 10 randomly sampled values.

```{r}
rnd <- replicate(1000, rnorm(n = 10, mean = 5, sd = 1))
ncol(rnd)
nrow(rnd)
```

We could get the sample means from each column using the command `colMeans()` and plot a histogram.

```{r}
xbar <- colMeans(rnd)
hist(xbar, breaks = 50, freq = FALSE)
```

This is a histogram of the sample means from each of our random samples. 

c) *Does it look like a normal distribution?*

We know from lectures that each sample mean will itself be a random sample from 
$$
\bar{x} \sim \mathcal{N}(\mu, \frac{\sigma}{\sqrt{n}})
$$

In our example, we have set $\mu = 5$, $\sigma = 1$ and $n = 10$.
Let's overlay this distribution to check.

```{r}
x <- seq(3, 8, length.out = 100)
y <- dnorm(x, mean = 5, sd = 1 / sqrt(10))
hist(xbar, breaks = 50, freq = FALSE)
lines(x, y)
```

We can clearly see that each sample mean is truly a single random sample from this distribution.

### Q5

In the above, we knew the variance ($\sigma = 1$) because we set this in our random sampling step.
In reality we don't know this, so we would use our sample variance ($s^2$) and a $t$-test.
Let's have a look at how the sample variances are distributed.

```{r}
library(matrixStats)
s2 <- colVars(rnd)
hist(s2, breaks = 50)
```

Notice that the sample variances ($s^2$) are not normally distributed around the true variance ($\sigma = 1$).

For these random samples, we knew the true value for the population mean  ($\mu = 5$).
In reality, **we wouldn't know this** and we'd need to conduct a hypothesis test with a null hypothesis and alternate hypothesis.

Let's test the hypothesis using $\mu = 5$ as the value of interest:

$$
H_0: \mu = 5 \quad \text{Vs} \quad H_A: \mu \neq 5
$$


To do this, we first calculate our $T$-statistic using:
$$
T = \frac{\bar{x} - \mu}{s / \sqrt{n}}
$$
Let's just do this on the first column to start with.

```{r}
Tstat <- (mean(rnd[,1]) - 5) / (sd(rnd[,1]) / sqrt(10))
df <- 10 - 1
```

NB: Here the degrees of freedom is n - 1, so we would check this against the distribution for $t_9$.
In my simulations I got the value $T =$ `r pander(Tstat)`.
**What value did you get from the first column in your simulations?**

Let's visualise this by first plotting the density of a $t_9$ distribution, then adding vertical lines for our obtained statistic.
By my estimation, 1 in 20 of you should get a relatively extreme $T$-statistic.
(Feel free to just copy & paste this code chunk, but do try to understand what each line is doing.)

```{r}
library(tidyverse)
x <- seq(-4, 4, length.out = 101) # Create a vector for the x-axis
y <- dt(x, df) # Find the value of the density at each value of x
data <- data_frame(x = x, y = y) # Combine x & y into a data_frame
ggplot(data, aes(x = x, y = y)) +
  geom_line() +
  geom_ribbon(data = filter(data, x > abs(Tstat)),
              aes(x= x, ymax =y, ymin = 0),
              fill = "grey") +
  geom_ribbon(data = filter(data, x < -abs(Tstat)),
              aes(x= x, ymax =y, ymin = 0),
              fill = "grey") +
  geom_vline(xintercept = c(-1, 1)*Tstat) +
  theme_bw() +
  labs(x = "T", y = "") +
  scale_y_continuous(expand = c(0, 0)) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        axis.line.x = element_line(colour = "black"),
        panel.border = element_blank(),
        panel.grid = element_blank())
```

a) *Do you think we have a low or high probability of observing this $t$-statistic?*
Let's check

```{r}
2*pt(abs(Tstat), df, lower.tail = FALSE)
```


# Multiple Hypothesis Testing

## Recap

We also learned about:

1. Type I and Type II errors
2. The Family-Wise Error Rate (FWER)
3. The Bonferroni Adjustment
4. The False Discovery Rate

### Q6

We could do this for every column in our matrix of random samples, which **we know** is a sample of 10 from $\mathcal{N}(5, 1)$

```{r}
Tstat <- (colMeans(rnd) - 5) / (colSds(rnd) / sqrt(10))
pValues <- 2*pt(abs(Tstat), df, lower.tail = FALSE)
```

In many analyses, we use $p < 0.05$ to reject $H_0$

a) *How many pValues out of our $m = 1000$ tests do we see below 0.05?*

```{r}
sum(pValues < 0.05)
```

b) Should we reject $H_0$ for these random samples? 
c) If so, would this be a correct rejection of $H_0$, a Type I error or a Type II error?

We could use the Bonferroni procedure to reduce our Type I errors, and control the Family Wise Error Rate (FWER)

```{r}
alpha <- 0.05/1000
sum(pValues < alpha)
```

In `R` we can simply adjust our $p$-values using the function `p.adjust()`.
This will give identical results to the above, but instead each adjusted-$p$ value will be $min(1, p*m)$, where $m = 1000$ and represents the number of tests we are performing.

```{r}
adjustedP <- p.adjust(pValues, method = "bonferroni")
sum(adjustedP < 0.05)
```

An alternative we briefly mentioned in lectures would be to allow a small number of false discoveries in our set of results.
This is known as the FDR and we can easily obtain this in `R`.

```{r}
fdrP <- p.adjust(pValues, method = "fdr")
sum(fdrP < 0.05) 
```

d) *Did you obtain different results using the two different methods?*
e) *Explain why or why not?*

# Non-parametric Testing

## The Wilcoxon Rank-Sum Test

All of the above examples refer to normally distributed data.
Let's say we have some pretty extreme values.

```{r}
sample1 <- c(5.5, 2.45, 1000, 6)
sample2 <- c(-1.1, 1.2, 2.4, -40)
```

### Q7

a) *Do you think these values come from the same distribution?*

Let's do a quick $t$-test in `R` and see what the results are.
The null and alternative hypotheses for a $t$-test are:
$$
H_0: \mu_1 = \mu_2 \quad \text{Vs} \quad H_A: \mu_1 \neq \mu_2
$$

```{r}
t.test(sample1, sample2)
```

We can clearly see that these two samples are drawn from different distributions.
The problem is that our data is not normally distributed (especially sample 1) so the $t$-test cannot cope with these extreme violations of normality.
It is also pretty clear that for data like this, our estimates of the population variance will be much larger than the difference.

An alternative would be the Wilcoxon Rank-Sum Test, where each value is given a rank and the ranks are compared.

$$
H_0: \text{The two samples are drawn from the same distribution} \\
VS  \\
H_A: \text{The two samples are drawn from different distributions}
$$

```{r}
wTest <- wilcox.test(sample1, sample2)
wTest
```

The ranks for each value can be seen using the following code:

```{r}
 order(c(sample1, sample2), decreasing = TRUE)
```

Note that because we use the ranks, instead of the values, this procedure is robust to changes in the actual values which don't affect the ranks.

b) *If we change the value `1000` to `10` will it change the results of the Wilcoxon Rank Sum Test?*
c) *Would it change the results for a $t$-test?*

```{r}
sample1[3] <- 10
sample1
t.test(sample1, sample2)
wilcox.test(sample1, sample2)
```

## Fisher's Exact Test

Fisher's Exact Test can be used for comparing counts in a $2 \times 2$ contingency table.
In the hypoxia example above these values would be:

```{r}
counts <- matrix(c(10, 94, 32, 86), ncol = 2)
rownames(counts) <- c("Hypoxia", "Normoxia")
colnames(counts) <- c("Allele A", "Allele B")
```

```{r, echo=FALSE, results='asis'}
pander(counts, justify = c("lrr"))
```

Here the null hypothesis & alternative are:

$$
H_0: \text{There is no association between alleles and survival in hypoxic conditions} \\
Vs \\
H_A: \text{There is an association between alleles and survival in hypoxic conditions}
$$

### Q8 

a) *Do you think there is an association just by looking at the data?*

In `R` we can simply conduct the test using `fisher.test()`

```{r}
fisher.test(counts)
```

Alternatively, we could've used a $\chi^2$ test, however we would've need to ensure that no cell in the table had an expected count $<5$.

```{r}
chisq.test(counts)
```

b) *Would you reject $H_0$ using either test?*

