---
title: "Using Data Frames"
author: "Steve Pederson"
date: "5 September 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = 'hide')
```

```{r}
library(knitr)
```

# Creating our own data frame

We can create a `data.frame` manually as well as loading from a file.
Let's create two vectors for some experimental mice.
If you still have the vector `lifeSpans` in your workspace, you don't need to recreate this.
```{r}
lifeSpans <- c(2.1, 1.9, 1.7, 1.8, 2.3, 2.5)
Groups <- c("treated", "control", "control", "control", "treated", "treated")
Groups <- factor(Groups)
```

Now we can combine these into a `data.frame`
```{r}
mouseLife <- data.frame(lifeSpans, Groups)
```

Note how R automatically named the columns.
We could also specify these manually.
 
```{r}
mouseLife <- data.frame(lifeSpans, Groups,
                        sex  = c("M", "M", "M", "F", "F", "F"))
mouseLife
```

We can easily subset a `data.frame` based on values in a column.

```{r}
subset(mouseLife, sex =="M")
```

We can also call a column individually, using the `$` symbol followed by the column name.

```{r}
mouseLife$Groups
```

## A Few T-Tests

With our `data.frame` we can conduct some $t$-tests.

### Using all of our data

First we we could test the null hypothesis:

$$
H_0: \mu = 2  
$$

with alternative hypothesis

$$
H_A: \mu \neq 2
$$

where $\mu$ is the true mean lifespan

**Here we'll need to just call the column of `mouseLife` called `lifeSpans`, which we can perform using the `$` symbol**

```{r, eval=FALSE}
t.test(mouseLife$lifeSpans, mu = 2)
```

*Note that we have now given the function `t.test` an extra parameter called `mu`*

If we didn't specify `mu = 2`, the function defaults to `mu = 0`.

### Testing `control` Vs `treated`

To perform this test, we can use the *R* formula syntax, where the symbol `~` can be interpreted as:
- depends on, or
- as a function of

```{r, eval=FALSE}
t.test(lifeSpans~Groups, mouseLife)
```

### Getting Help

To call the help page on any function, we just preface it with a `?`

```{r, eval=FALSE}
?t.test
```

Sometimes *R* help pages can be tricky to understand.

In our code above we have used the second and third versions of the function described on this page.
*R* has automatically detected which version to call by our placement of either a vector in the first position (`mouseLife$lifeSpans`), or by the placement of a formula (`lifeSpans~Groups`).

- In the first test, we specified the additional argument `mu` by name.
- In the second test, we specified the object `mouseLife` as our data object by placing it after the formula.


